---
date: '2023-08-04'
output:
  html_document:
    keep_md: yes
---

### Population Stratification
**Developed by Juliana Acosta-Uribe for the ReD-Lat Consortium 2023**

#### Data:
In this tutorial we will perform a Principal Component Analysis (PCA) of a subset of individuals from the [1000 Genome project](https://www.internationalgenome.org/) `1000G.toy`

However, you can use your own plink formatted .*bed*, .*fam*, .*bim* set
 If you are analyzing more than one population, you will have to provide an additional file `sample_data.txt` specifying which population does each samples belong to.
 Header should be as follows: \
 **IID** ID of the sample as it is in the plink IID or in the VCF
 **cluster_column** It can be any name or number to differentiate the different populations. e.g. in my data it coprresponds to the country of origin

R is expecting a tab delimited file. If your file is delimited by spaces you can fix it with the following bash command `sed -i 's/ /\t/g'  sample_data.txt`


#### Software:
-[R](https://www.r-project.org/) or [RStudio](https://posit.co/download/rstudio-desktop/)\
-[plink](https://www.cog-genomics.org/plink2/)\
-[king](https://www.kingrelatedness.com/) \
-[ADMIXTURE](https://dalexander.github.io/admixture/download.html) \
Alexander, David H., and Kenneth Lange. "Enhancements to the ADMIXTURE algorithm for individual ancestry estimation." BMC bioinformatics 12 (2011): 1-6.\
-[pong](https://github.com/ramachandran-lab/pong) \
Aaron A. Behr and others, pong: fast analysis and visualization of latent clusters in population genetic data, Bioinformatics, Volume 32, Issue 18, September 2016, Pages 2817–2823
-[EIGENSOFT](https://github.com/DReichLab/EIG) \
Patterson et al. 2006 PLoS Genet 2:e190 (population structure)


#### Environment 

This script is designed to be run as an [R markdown](https://rmarkdown.rstudio.com/lesson-1.html) file in R Studio. \
This way you can run it in a step-by-step mode. You could also run it directly from the r command line if you already have a dataset of unrelated individuals and a dataframe specifying which population they belong to. 
```         
library(rmarkdown)
render("path/to/your/file.Rmd")
```

Specify your prefix and the path for your working directory and required software in the following chunk:
```{r environment-setup}
if (!require("knitr", quietly = TRUE))
install.packages("knitr")
library(knitr)

# Set your working directory:
## Edit 'root.dir' and 'setwd' with the path to your working directory
knitr::opts_chunk$set(root.dir = "~/Redlat/PCA",
                      dev = "png",
                      dpi = 300,
                      echo = FALSE,
                      cache = TRUE)
setwd("~/Redlat/PCA")
date = Sys.Date() # gets the current date

# Set up path to software:
## Give the paths to plink, king
Sys.setenv(plink='/home/acostauribe/bin/plink')
Sys.setenv(king='/home/acostauribe/bin/king')
Sys.setenv(admixture='/home/acostauribe/bin/admixture')
Sys.setenv(pong='/home/acostauribe/anaconda3/bin/pong')
Sys.setenv(smartpca='/home/acostauribe/bin/EIG-8.0.0/bin/smartpca.perl')

# Give the name of your starting file without the .bed or .vcf.gz extension
prefix='redlat_merge.id.kin.concordant'
Sys.setenv(prefix=prefix)

#Set a color palette that is unambiguous both to colorblinds and non-colorblinds
col_palette = c( '#4477AA', '#EE6677', '#228833', '#CCBB44', '#66CCEE', '#A9A9A9', '#AA3377')

```


### Principal Component Analysis

**1. Get a set of unrelated Samples** 

We will use King to output a set of unrelated individuals from our cohort.
This will create a list of unrelated individuals (prefix.king.unrelated.txt) that you can extract from using plink
```{bash get-unrelated, eval=FALSE, include=FALSE}
$king -b $prefix.bed --unrelated --prefix $prefix.king.
$plink --bfile $prefix --keep $prefix.king.unrelated.txt --make-bed --out $prefix.unr
```

**2. Retain variants with Minor Allelic Frequency > 10%**

If you have a balanced dataset, with similar number of individuals from each included population, you can filter your allelic frequency based on the markers in your dataset.

```{bash filter-maf, eval=FALSE, include=FALSE}
$plink --bfile $prefix.unr --maf 0.10 --make-bed --out $prefix.unr.maf
```

If your set is unbalanced, with a population outnumbering the other, I recommend to use population based allelic frequencies from a balanced study to filter your samples. For the ReDLat consortium data, we used the global allelic frequencies from the 1000 Genomes Project to annotate our data and then we retained variants with maf > 10% according to those values. 

**3. Filter for Linkage Disequilibrium and retain independent markers**

Plink uses a *sliding window* approach to identify variants in linkage disequilibrium. There are many options to modify the behavior or this approach in [plink's docummentation](https://www.cog-genomics.org/plink/1.9/ld#indep). The LD pruning requires that the *.bim* file has variant IDs in the second column. If no variants have been assigned, you could do a preliminary step using [--set-missing-var-ids](https://www.cog-genomics.org/plink/1.9/data#set_missing_var_ids).

For LD pruning we will use the following:
--indep-pairwise <window size>['kb'] <step size (variant ct)> <r^2 threshold>

```{bash filter-ld, eval=FALSE, include=FALSE}
# Identify variants in LD
$plink --bfile $prefix.unr.maf --indep-pairwise 50 10 0.2 

# Retain independent variants
$plink --bfile $prefix.unr.maf --extract plink.prune.in --make-bed --out $prefix.unr.maf.ld
```

After the LD pruning we should have a dataset for > 100.000 variants to identify population structure.


**4. Perform a PCA using Plink**

Plink  calculate the eigenvectors and eigenvalues using the the PCA algorithm from [GCTA](https://yanglab.westlake.edu.cn/software/gcta/#PCA). 

```{bash pca, eval=FALSE, include=FALSE}
$plink --bfile $prefix.unr.maf.ld --pca header tabs --out $prefix.pca
```

Eigenvectors are written to *prefix.unr.maf.ld.pca.eigenvec*, and top eigenvalues are written to *prefix.unr.maf.ld.pca.eigenval*. The eigenvectors represent the directions or components in the original variable space, while the eigenvalues quantify the amount of variance explained by each eigenvector.

The 'header' modifier adds a header line to the .eigenvec file(s), and the 'tabs' modifier makes the .eigenvec file(s) tab- instead of space-delimited.

**4. Perform a PCA using the EIGENSOFT package**

`$smartpca` calls the smartpca program (EIG-8.0.0/POPGEN/README). For this to work, the bin directory containing
smartpca MUST be in your path.

smartpca ignores individuals in the .fam file if they are marked as missing in the phenotypes column. 

The awk command provides a new .fam file that will automatically include all individuals.

```{bash}
awk '{print $1, $2, $3, $4, $5, 1}' $prefix.unr.maf.ld.fam > $prefix.unr.maf.ld.pca.fam

perl $smartpca 
-i $prefix.unr.maf.ld.bed     # genotype file 
-a $prefix.unr.maf.ld.bim     # snp file
-b $prefix.unr.maf.ld.pca.fam     # indiv file
-e $prefix.pca.eval           # output file of all eigenvalues
-o $prefix.pca                # output file of principal components                
-l $prefix.smartpca.log       # output logfile  
-p $prefix.pca.plot
-m 0                          # dont remove outliers for the exploratory PCAls

```

`-m maxiter` (Default is 5) maximum number of outlier removal iterations. To turn off outlier removal, set -m 0. \
`-t topk` (Default is 10) number of principal components along which to remove outliers during each outlier 
removal iteration. \
`-s sigma` (Default is 6.0) number of standard deviations which an individual must exceed, along one of top k top
principal components, in order to be removed as an outlier. \

Eigenvectors are written to *prefix.unr.maf.ld.pca.evec*, and top eigenvalues are written to *prefix.unr.maf.ld.pca.eval*. The eigenvectors represent the directions or components in the original variable space, while the eigenvalues quantify the amount of variance explained by each eigenvector.


**5.Plot the PCA**

We will use ggplot2 to get a beautiful PCA plot
Plink's `prefix.unr.maf.ld.pca.eigenvec.eigenvec` contains the ei

Load the dataframe we are going to plot: \

If you used **plink** load using this chunk:
Notice that in our  previous command we determined the output to be separates with tab '\t', and to have a header
```{r load-plink, eval=FALSE, include=FALSE}
# Import data
eigenvec = read.delim(paste0(prefix,".pca.eigenvec"), sep='\t', header=TRUE)
eigenval = read.delim(paste0(prefix,".pca.eigenval"), sep='\t', header=FALSE)
```

If you used **Eigensoft** load using this chunk:
```{r load-eigensoft}
if (!require("tidyr", quietly = TRUE))
install.packages("tidyr")
library(tidyr)

# Import data
eigenvec = read.delim(paste0(prefix,".pca.evec"), sep='', header=FALSE)
eigenval = read.delim(paste0(prefix,".pca.eval"), header=FALSE)

# Edit header
eigenvec = eigenvec[-1, ] # Remove first line
pc_names = paste0("PC", 1:(ncol(eigenvec) - 1)) # Count the number of columns
names(eigenvec) =  c("FID:IID", pc_names) # rename columns
eigenvec = separate(eigenvec, "FID:IID", into = c("FID", "IID"), sep = ":") # Separate the family and individual IDs
```

We also need to add the correspondent country or site that each individual belongs to. For the data we will use the `sample_data.txt` dataframe. Using the `match` function, we will create a new column in 'eigenvec'.

```{r load-sample-data}
lookup_dataframe = read.delim("sample_data.txt", header = TRUE )

# Determine which Column you want to use as comparison factor
cluster_column = "country_birth"  # This can be any column name you want to use
eigenvec[[cluster_column]] = lookup_dataframe[[cluster_column]][match(eigenvec$IID, lookup_dataframe$IID)]
```

Plot 
```{r pca-scatter-plots}
if (!require("ggplot2", quietly = TRUE))
install.packages("ggplot2")
library(ggplot2)

if (!require("rlang", quietly = TRUE))
install.packages("rlang")
library(rlang)

prefix = paste0(prefix, ".", cluster_column)

# Perform Scatter plots of main PCs
ggplot(eigenvec, aes(x=PC1, y=PC2, color=.data[[cluster_column]])) + 
        geom_point( size=0.8, alpha=0.7) +
        theme_linedraw() + 
        guides(color = guide_legend(override.aes = list(size = 5))) +
        scale_fill_manual(values=col_palette) +
        scale_colour_manual(name=NULL, values=col_palette) +
        labs(title="Principal Component Analysis PC1 vs PC2",
             caption=date)
ggsave(paste0(prefix,".PC1-PC2.png"), width = 6, height = 5, dpi = 300)

ggplot(eigenvec, aes(x=PC2, y=PC3, color=.data[[cluster_column]])) + 
        geom_point( size=0.8, alpha=0.7) +
        theme_linedraw() + 
        guides(color = guide_legend(override.aes = list(size = 5))) +
        scale_fill_manual(values=col_palette) +
        scale_colour_manual(name=NULL, values=col_palette) +
        labs(title="Principal Component Analysis PC2 vs PC3",
             caption=date)
ggsave(paste0(prefix,".PC2-PC3.png"), width = 6, height = 5, dpi = 300)

ggplot(eigenvec, aes(x=PC1, y=PC3, color=cluster_column)) + 
        geom_point( size=0.8, alpha=0.7) +
        theme_linedraw() + 
        guides(color = guide_legend(override.aes = list(size = 5))) +
        scale_fill_manual(values=col_palette) +
        scale_colour_manual(name=NULL, values=col_palette) +
        labs(title="Principal Component Analysis PC1 vs PC3",
             caption=date)
ggsave(paste0(prefix,".PC1-PC3.png"), width = 6, height = 5, dpi = 300)

```


Perform violin plot of each country vs top Principal Components

```{r pcs-according-population}
theme_2 = theme(panel.background = element_rect(fill=NA),
             legend.position="none",   
             panel.border=element_rect(fill=NA),
             plot.title = element_text(hjust = 0.5),
             axis.title.x = element_blank())

# State the number of principal components you want to plot
pcs = 5

plot_pc = paste0("PC", 1:(pcs)) # Count the number of columns

for (pc in plot_pc) {
    # Create the plot for the current PC
    p = ggplot(eigenvec, aes(x=.data[[cluster_column]],
                             y=.data[[pc]], 
                             color=.data[[cluster_column]], 
                             fill=.data[[cluster_column]])) + 
        geom_violin(trim=FALSE, alpha=0.9) +
        geom_boxplot(width=0.1, color="black", outlier.size=0.5, alpha=0.9) +
        theme_2 + 
        scale_colour_manual(values=col_palette) +
        scale_fill_manual(values=col_palette) +
        labs(title=paste("RedLat", pc, "according to", cluster_column),
             caption=date)
    
    # Print the plot
    print(p)

    # Save the plot
    filename = paste(prefix, pc, cluster_column, "png", sep = ".")
    ggsave(filename, plot = p, width = 6, height = 5, dpi = 300)
}


```
The following code generates a pair-plot our top 4 principle components. As ggplot does not have this functionality, we have to use base R for plotting

```{r}
# Assign a Color to each country
col_countries= c("Argentina" = '#4477AA', "Brazil" = '#EE6677', "Chile" = '#228833', "Colombia" = '#CCBB44', "Mexico" = '#66CCEE', "Other" = '#A9A9A9', "Peru" = '#AA3377')

# Start PNG device
png(paste0(prefix, ".PCA_pairs_plot.png"), width = 2400, height = 2400, res = 300)

# Create a pair plot with scatter plots
pairs(eigenvec[, c("PC1", "PC2", "PC3", "PC4")],
      pch = 19,
      col = alpha(col_countries[eigenvec$cluster_column], 0.4),
      main = "Principal Component Analysis")
mtext(date, side = 1, line =4, cex = 0.5)

# Close PNG device
dev.off()

# Plot a legend for labels and colors
png(paste0(prefix, ".PCA_pairs_plot_legend.png"), width = 2400, height = 800, res = 300)

plot(NULL,xaxt='n',yaxt='n',bty='n',ylab='',xlab='', xlim=0:1, ylim=0:1)
legend_plot = legend("top", legend = names(col_countries), col = col_countries, pch = 19, title = "Sequencing type",
                      horiz = TRUE, cex = 0.7)
dev.off()
```

Perform a Scree plot to check the explained variance
```{r variance-expained-pc}
#Edit the eigenvalues and add the corresponded PC to each value
colnames(eigenval) = "values"
eigenval$PC = 1:nrow(eigenval)

# Select the number of Eigenvalues you want to plot
eigenval_subset = 10
eigenval_plot = eigenval[1:eigenval_subset, ]

# Sum all eigenvalues in the original file
total_sum = sum(eigenval$values)

# Identify the proportion explianed by each of PC that will be plotted
eigenval_plot$proportions = (eigenval_plot$values / total_sum)

ggplot(eigenval_plot, 
       aes(x=PC, y=proportions, color='4477AA', fill='4477AA' )) + 
       geom_bar(stat="identity") +
       theme_2 + 
        scale_fill_manual(values=col_palette) +
        scale_colour_manual(values=col_palette) +
        scale_x_continuous(name="Principal Component", breaks=seq(1,eigenval_subset,1)) +
        scale_y_continuous(name="Variance Proportion") +
        labs(title=paste("Variance explained by top", eigenval_subset, "Principal Components", sep = " "),
             caption=date)
ggsave(paste0(prefix, eigenval_subset, ".pc-scree-plot.png"), width = 6, height = 5, dpi = 300)
```

## ADMIXTURE Analysis

Choose number of iterations and populations (K) you want to do:

```{r set-up admixture}
admixture_iterations=20
Sys.setenv(admixture_iterations=admixture_iterations)

k_model=10
Sys.setenv(k_model=k_model)
```

Run admixture in an unsupervised mode
```{bash run-admixture, eval=FALSE, include=FALSE}

declare -i iterations=$admixture_iterations
declare -i model=$k_model

for i in $(seq 1 $iterations)
do
        echo "Starting" Iteration_${i}
  	mkdir Iteration_${i}
        for k in $(seq 1 $model) 
        do
                echo "Starting K of " ${k}
                $admixture -j4 $prefix.unr.maf.ld.bed --cv ${k} -s time | tee $prefix.unr.maf.ld.K${k}.I${i}.log
                #creates $prefix.unr.maf.ld.${k}.P, $PREFIX.unr.maf.ld.${k}.Q and $prefix.unr.maf.ld.K${k}.I${i}.log
                
                mv $prefix.unr.maf.ld.K${k}.* ./Iteration_${i}
                mv *.P ./Iteration_${i}
                mv *.Q ./Iteration_${i}
                echo "Finished K of " ${k}
        done
        echo "Finished" Iteration_${i}
done
```
Each run of admixture will create the following files:
- *prefix.unr.maf.ld.{k}.P* each line is a variant
- *prefix.unr.maf.ld.{k}.Q *
- *prefix.unr.maf.ld.K{k}.I{i}.log*


**Get the Cross Validation (CV) error from each run**

We can use a bash function called 'grep' to create a single file with all the CV errors from the .log files
```{bash cv-error}
declare -i iterations=$admixture_iterations
declare -i model=$k_model

for i in $(seq 1 $iterations)
do
        echo "Starting" Iteration_${i}
        for k in $(seq 1 $model) # choose number of ancestral populations (K) you want to model
        do
                echo "Starting K of " ${k}
                grep "CV error" ./Iteration_${i}/$prefix.unr.maf.ld.K${k}.I${i}.log >> $prefix.CVerror.txt
        done
done
```

We can use a bash function called 'sed' to edit $prefix.CVerror.txt.
``sed -i 's/STRING_TO_REPLACE/STRING_TO_REPLACE_IT/g' filename``

```{bash edit-cv-error, eval=FALSE, include=FALSE}
sed -i 's/CV error (K=//g' $prefix.CVerror.txt
sed -i 's/)://g' $prefix.CVerror.txt
```

**Plot the CV errors for each K**

```{r plot-cv-error}
if (!require("ggplot2", quietly = TRUE))
install.packages("ggplot2")
library(ggplot2)

# Import data
admixture_data = read.delim(paste0(prefix,".CVerror.txt"), header =F, sep=" ")

# Add Column names
colnames(admixture_data) = c("K", "CV.error")

# Get the mean of each CV error per K
mean_cv = aggregate(admixture_data, list(admixture_data$K), mean)

# Plot 
ggplot(admixture_data, aes(x = K, y = CV.error)) +
  geom_point(shape = 16, color = "blue", alpha = 0.3) +
  xlab("Ancestral populations (K)") +
  ylab("CV error") +
  ggtitle("Cross Validation Error") +
  theme_bw() +
  geom_line(data = mean_cv, aes(x = K, y = CV.error), color = "red", linetype = "dashed") +
  scale_x_discrete(limits=c(seq(1:k_model)))
ggsave(filename = "cverror.png",
       width = 6, height = 4, dpi = 300)
```


**Plot your ADMIXTURE results using pong**

You can use base R or ggplot to plot your ADMIXTURE results, but i recommend a software called pong. \
Pong installation will not be covered by this tutorial. More information on the Ramachandran Lab [GitHub page](https://github.com/ramachandran-lab/pong)

pong accepts clustering output files, or Q matrices, from a variety of clustering inference programs, including ADMIXTURE

These are the arguments we will be using for our plot:
``-m FILEMAP, --filemap FILEMAP``

A FILEMAP is a three-column, tab-delimited file. Each line contains the following information about a single Q matrix:

- A unique ID (e.g. iter5_K=7; this must contain at least one letter, and cannot contain a hashmark/hashtag or a period. Integers, spaces, and other special characters are accepted.)
- The K value
- The Q matrix file path, relative to the filemap. Thus, if the filemap is in the same directory as the Q matrix files, this is just the name of the Q matrix file.

It is important that the columns of your filemap are in the right order and that the file is tab-delimited. Use the filemap provided with the example dataset for reference. U

example:
```
I1_K2	2	/Iteration_1/1000G.toy.unr.maf.ld.2.Q
I1_K3	3	/Iteration_1/1000G.toy.unr.maf.ld.3.Q
I2_K2	2	/Iteration_2/1000G.toy.unr.maf.ld.2.Q
I2_K3	3	/Iteration_2/1000G.toy.unr.maf.ld.3.Q
I3_K2	2	/Iteration_3/1000G.toy.unr.maf.ld.2.Q
I3_K3	3	/Iteration_3/1000G.toy.unr.maf.ld.3.Q
```

You can generate your 'filemap' using this bash script
```{bash cv-error-pong, eval=FALSE, include=FALSE}
declare -i iterations=$admixture_iterations
declare -i model=$k_model

for i in $(seq 1 $iterations)
do
        for k in $(seq 1 $model)
        do
                echo -e "I"$i"_K"$k"\t"$k"\t./Iteration_"${i}"/"$prefix".unr.maf.ld."$k".Q" >> filemap.txt
        done
done
```


``-i IND2POP, --ind2pop IND2POP`` \
ind2pop data (can be either a Q-matrix column number or the path to a file containing the data).

Each line of the Q matrix file represents one individual, and it has the same order as the .fam file.
Each row is the population that the corresponding individual belongs to.
                        
``-n POP_NAMES, --pop_names POP_NAMES`` \
Path to file containing population order/names.


```{bash pong, eval=FALSE, include=FALSE}
$pong --filemap filemap.txt --ind2pop ind2pop.txt --pop_names pop_names.txt 
```
